---
title: "Additional examples"
subtitle: "Vignette 3 of 4"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: TRUE
    number_sections: TRUE
    toc_depth: 3
    fig_caption: yes
    fig_width: 9
    fig_height: 6
---

<!-- badges: start -->
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/precmed)](https://cran.r-project.org/package=precmed)
[![metacran downloads](https://cranlogs.r-pkg.org/badges/last-month/precmed)](https://cran.r-project.org/package=precmed)
<!-- badges: end -->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
options(rmarkdown.html_vignette.check_title = FALSE) #title of doc does not match vignette title
doc.cache <- T #for cran; change to F
```


<h1>precmed: Precision Medicine in R</h1>
A doubly robust precision medicine approach to estimate and validate conditional average treatment effects

# Additional examples for the `precmed` package
**precmed** can be flexible given the many arguments that the user can specify. The example described above is one of the many ways to perform the analysis with most argument values kept as default. We provide a few more examples below to show what other options the user has and how to be more creative in using **precmed**. 

# Load required packages
```{r, echo = T, message=F}
library(precmed)
library(dplyr)
library(ggplot2)
```

# Additional examples with count outcomes
## Initial predictor method
If you choose two regressions or contrast regression as the scoring method, there is an option to specify how the initial outcome regression estimates ($\hat\mu_r^k$ where $r = 0,1$ is the treatment and $k$ is the CV fold) are calculated. See [Theoretical details](Theoretical-details.html) section for the background of these initial predictors and how they are used in two and contrast regressions. Here is a list of all 3 options:

* "poisson" (Poisson regression from the `glm()` function) 
  * generalized linear model, strong assumption on Poisson distribution, fast
* "gbm" (gradient boosting machine from the `gbm` R package)
  * ensemble method, typically tree-based, slow
* "gam" (generalized additive model from the `mgcv` package)
  * a combination of generalized linear model and additive models where the linear predictor is an additive function of the covariates, can be slow

The main example above used a Poisson regression but you can try the other 2 non-linear and more flexible options. Below, we use `catecvcount()` as the example code but this can be applied to `catefitcount()` as well. 

```{r initial_predictor_gbm_gam_count, eval = F, echo = T, cache = doc.cache}
# An example of using GBM as the initial predictor method
catecv(response = "count",
       data = countExample,
       score.method = c("poisson", "boosting", "twoReg", "contrastReg", "negBin"),
       cate.model = y ~ age + female + previous_treatment + previous_cost + previous_number_relapses + offset(log(years)),
       ps.model = trt ~ age + previous_treatment, 
       higher.y = FALSE,
       initial.predictor.method = "boosting",       # NEW
       cv.n = 5, 
       seed = 999,
       plot.gbmperf = FALSE,
       verbose = 0)

# An example of using GAM as the initial predictor method
catecv(response = "count",
       data = countExample,
       score.method = c("poisson", "boosting", "twoReg", "contrastReg", "negBin"),
       cate.model = y ~ age + female + previous_treatment + previous_cost + previous_number_relapses + offset(log(years)),
       ps.model = trt ~ age + previous_treatment, 
       higher.y = FALSE, 
       initial.predictor.method = "gam",        # NEW
       xvar.smooth = c("age", "previous_cost"), # NEW
       cv.n = 5,
       seed = 999,
       plot.gbmperf = FALSE,
       verbose = 0)
```

Note that for the GAM method, there is an additional argument called `xvar.smooth`. If left as default (which is NULL), GAM will include all independent variables in `cate.model` as smooth terms with function `s()` in the linear predictor. You can choose to specify only a subset of the dependent variables as the smooth terms. In the example above, the specifications `cate.model` and `xvar.smooth` lead to the following GAM formula: `y ~ female + previous_treatment + previous_number_relapses + s(age) + s(previous_cost)`, where only age and previous medical costs are wrapped in the smooth function and the remaining 3 independent variables are not. 

Generally speaking, the "poisson" initial predictor method is the fastest and "gbm" is the slowest among the 3 methods, while "gam" is somewhere in between with longer computational time as the number of smooth terms increases.


## Subgroup proportion
The ATE by subgroups and validation curve results depend on how the subgroups are defined, which is determined by the sorted CATE scores. There could be nested binary subgroups (cutoffs specified by `prop.cutoff`) or mutually exclusive subgroups (cutoffs specified by `prop.multi`). In the main example above, we used the default values of `prop.cutoff` and `prop.multi` and we have provided explanations in the ATE results. Here is a recap:

* `prop.cutoff = seq(0.5, 1, length = 6)` means that we have 6 sets of nested binary subgroups where the first subgroup splits the sample by 50/50 according to the estimated CATE scores (sorted), the 2nd subgroup splits the sample by 60/40, ..., the 5th subgroup splits the sample by 90/10, and the 6th subgroup splits the sample by 100/0. The validation curves show the ATE in the first group of each split, i.e., 50%, 60%, 70%, ..., 100%. The 6th subgroup is technically not a split but we keep it because it is equivalent to the overall ATE (the gray dashed reference line). The functions that use `prop.cutoff` will automatically discard 0 if supplied because 0/100 split does not make sense as there is no 0% subgroup to plot for the validation curves, and a warning message will be given. The argument `higher.y` controls the direction of the sorted CATE scores, i.e., whether we split from the lowest or the highest CATE scores. For our toy example, `higher.y` = FALSE so we split from the lowest CATE scores such that the first group in each subgroup represents higher responders to the treatment coded with 1.

* `prop.multi = c(0, 1/3, 2/3, 1)` means that we specify a 3-category mutually exclusive subgroup, split by 33/33/33 according to the estimated CATE scores. The argument `higher.y` controls the direction of the sorted CATE scores, i.e., where we split from the lowest or the highest CATE scores. For our toy example, `higher.y` = FALSE so we split from the lowest CATE scores. Contrary to `prop.cutoff`, `prop.multi` must include both 0 and 1 to remind us of the mutual exclusiveness. he functions that use `prop.multi` will automatically attach 0 and/or 1 if they are not supplied and a warning message will be given. 
Now we show more examples of different cutoffs and how they are reflected in the results. 

```{r subgroup_proportion_count, eval = T, echo = T, cache = doc.cache}
# An example of 11 nested binary subgroups and a 4-category mutually exclusive subgroup
output_catecv2 <- catecv(response = "count",
                         data = countExample,
                         score.method = c("poisson", "contrastReg", "negBin"),
                         cate.model = y ~ age + female + previous_treatment + previous_cost + previous_number_relapses + offset(log(years)),
                         ps.model = trt ~ age + previous_treatment,
                         higher.y = FALSE,
                         prop.cutoff = seq(0.5, 1, length = 11), # NEW
                         prop.multi = c(0, 1/4, 2/4, 3/4, 1),    # NEW
                         initial.predictor.method = "poisson", 
                         cv.n = 5, 
                         seed = 999,
                         plot.gbmperf = FALSE,
                         verbose = 0)
```

```{r subgroup_proportion2_count, eval = T, echo = T}
print(output_catecv2$ate.contrastReg$ate.est.train.high.cv) 
# Dimension is now 11 rows (nested binary subgroups) by 5 columns (CV iterations)
```

Because we have new `prop.cutoff` with 11 values as opposed to only 6 in the original example, the number of rows in the ATE results will be changed accordingly while the number of columns stays the same because we still use 5 CV iterations. 


```{r subgroup_proportion3_count, eval = T, echo = T}
plot(output_catecv2) 
```

The validation curves depend on `prop.cutoff` and are now smoother than the one presented in the main example because they contain 10 subgroups instead of 5. The ABC statistic is not affected by the change in `prop.cutoff` because the calculation of ABC statistics depends on the minimum and maximum of the `prop.cutoff` vector and is always calculated based on 100 evenly-spaced cutoffs created from this range. See more theoretical details on ABC calculation in the [Theoretical details](Theoretical-details.html) section. 


```{r subgroup_proportion4_count, eval = T, echo = T}
print(output_catecv2$ate.contrastReg$ate.est.train.group.cv) 
# Dimension is now 4 rows (multi-category mutually exclusive subgroups) by 5 columns (CV iterations)

boxplot(output_catecv2)
```

For results on the mutually exclusive subgroups, the number of rows in the ATE results increases from 3 to 4 in comparison with the original example because `prop.multi` now defines 4 mutually exclusive subgroups. For each method, the box plot now has 4 boxes for 4 mutually exclusive subgroups.


**Examples that we do not recommend**

In this next example, `prop.cutoff` includes 0.01, which is very close to 0. Consequently, the scoring methods may run into some convergence issues because the first subgroup includes a small number of patients. This can lead to highly unstable ATE estimates, as shown in the first CV iteration of contrast regression in `output_catecv3$ate.contrastReg$ate.est.valid.high.cv`. Moreover, the validation curves become useless due to the extreme ATE estimates.

```{r subgroup_proportion_bad_count, eval = T, echo = T, cache = doc.cache}
# An example of very few nested binary subgroups
output_catecv3 <- catecv(response = "count",
                         data = countExample,
                         score.method = c("poisson", "contrastReg", "negBin"),
                         cate.model = y ~ age + female + previous_treatment
                                      + previous_cost + previous_number_relapses 
                                      + offset(log(years)),
                         ps.model = trt ~ age + previous_treatment, 
                         initial.predictor.method = "poisson", 
                         higher.y = FALSE,
                         prop.cutoff = c(0, 0.01, 0.30, 0.75, 1), # NEW
                         cv.n = 5, 
                         seed = 999,
                         plot.gbmperf = FALSE,
                         verbose = 1)
output_catecv3$ate.contrastReg$ate.est.valid.high.cv
```

```{r subgroup_proportion_bad2_count, eval = T, echo = T, warning=FALSE, message=FALSE}
plot(output_catecv3) 
```

Below are some general recommendations on the choice of `prop.cutoff` and `prop.multi`:

1. We do not recommend including 0 in `prop.cutoff` as it is an invalid value. The functions will automatically remove 0 in `prop.cutoff` and output a warning message.
2. We do not recommend choosing values too close to 0 either because this will generate very small subgroups which may lead to numerical instability, e.g., 1/99 split specified by 0.01 in this example and corresponding ATE in the training data for contrast regression. In this situation, either the validation curves will be dominated by the unstable ATE estimates, making the plot useless as in this example, or the validation curves will not be plotted if the numerical instability due to an extreme split leads to missing values, in which case a warning is printed as in the example above. 
3. In general, we do not recommend specifying too few proportions for `prop.cutoff` (e.g., 2 or 3) because the validation curves will look jagged. The validation curves will be smoother with more proportions spread out over the range of proportions. If computation is an issue, however, choosing too many values for `prop.cutoff` or `prop.multi` can be time consuming. It is recommended to start from a smaller length and you can increase the length as needed. 


# Additional examples for survival outcomes


## Initial predictor method

If you choose two and/or contrast regression as the scoring method, there is an option to specify how the initial outcome regression estimates ($\hat\mu_r^k$ where $r = 0,1$ is the treatment and $k$ is the cross validation fold) are calculated. See [Theoretical details](Theoretical-details.html) section for the background of these initial predictors and how they are used in the two and contrast regression. Here is a list of all 3 options:

-   "logistic" (logistic regression from the `glm()` function)
    -   generalized linear model, strong assumption on binomial
        distribution, fast
-   "boosting" (gradient boosting machine from the `gbm` R package)
    -   ensemble method, typically tree-based, assume the time-to-event
        outcome as Gaussian distributed, slow
-   "randomForest" (random survival forest model from the
    `randomForestSRC` package)
    -   ensemble method, tree-based, non-parametric, can be slow

The main example above used a logistic regression but you can try the
other 2 non-linear and more flexible options. Below, we use `catecv()`
as the example code but this can be applied to `catefit()` as well.

```{r initial_predictor_gbm_gam, eval = F, echo = T, cache = doc.cache}
# An example of using GBM as the initial predictor method
tau0 <- with(survivalExample,
             min(quantile(y[trt == "drug1"], 0.95), quantile(y[trt == "drug0"], 0.95)))

catecv(response = "survival",
       data = survivalExample,
       score.method = c("poisson", "boosting", "randomForest", "twoReg", "contrastReg"),
       cate.model = survival::Surv(y, d) ~ age + female + previous_cost + previous_number_relapses,
       ps.model = trt ~ age + previous_treatment,
       ipcw.model = NULL,
       followup.time = NULL,
       tau0 = tau0,
       surv.min = 0.025,
       higher.y = TRUE,
       prop.cutoff = seq(0.6, 1, length = 5),
       prop.multi = c(0, 0.5, 0.6, 1),
       cv.n = 5,
       initial.predictor.method = "boosting", # NEW
       tree.depth = 3,                        # NEW
       seed = 999,
       plot.gbmperf = FALSE,
       verbose = 0)

# An example of using random forest as the initial predictor method
catecv(response = "survival",
       data = survivalExample,
       score.method = c("poisson", "boosting", "randomForest", "twoReg", "contrastReg"),
       cate.model = survival::Surv(y, d) ~ age + female + previous_cost + previous_number_relapses,
       ps.model = trt ~ age + previous_treatment,
       ipcw.model = NULL,
       followup.time = NULL,
       tau0 = tau0,
       surv.min = 0.025,
       higher.y = TRUE,
       prop.cutoff = seq(0.6, 1, length = 5),
       prop.multi = c(0, 0.5, 0.6, 1),
       cv.n = 5,
       initial.predictor.method = "randomForest", # NEW
       n.trees.rf = 500,                          # NEW
       seed = 999,
       plot.gbmperf = FALSE,
       verbose = 0)
```

Note that for `boosting` and `randomForest` methods that are tree-based, parameters such as tree depth and number of trees can be specified with arguments `tree.depth`, `n.trees.rf`, and `n.trees.boosting`.

Generally speaking, the "logistic" initial predictor method is the fastest and "boosting" is the slowest among the 3 methods, while "randomForest" is somewhere in between with longer computational time as the number of trees and tree depth increase.

## Subgroup proportion

The **precmed** results depend on the subgroups which are determined by the sorted CATE scores. There could be nested binary subgroups (cutoffs specified by `prop.cutoff`) or mutually exclusive subgroups (cutoffs specified by `prop.multi`). In the main example above, we used certain values of `prop.cutoff` and `prop.multi` and we have provided explanations in the ATE results. Here is a recap:

* `prop.cutoff = seq(0.6, 1, length = 5)` means that we have 5 sets of     nested binary subgroups where the first subgroup set is split by 60/40 according to the estimated CATE scores (sorted), the 2nd subgroup set is split by 70/30, ..., and the 6th subgroup set is split by 100/0. The validation curves show results of the first group of each split, i.e., 60%, 70%, ..., 100%. The 5th subgroup set is technically not a split but we keep it because it is equivalent to the overall ATE (the gray dashed reference line). **precmed** will automatically discard 0 if supplied because 0/100 split does not make sense as there is no 0% subgroup to plot for the validation curves, and a warning message will be given. Argument `higher.y` controls the direction of the sorted CATE scores, i.e., whether we split from the lowest or the highest CATE scores. For our toy example, `higher.y` = TRUE so we split from the lowest CATE scores.

* `prop.multi = c(0, 0.5, 0.6, 1)` means that we specify a 3-category mutually exclusive subgroup, split by 50/10/40 according to the estimated CATE scores. Argument `higher.y` controls the direction of the sorted CATE scores, i.e., where we split from the lowest or the highest CATE scores. For our toy example, `higher.y` = TRUE so we split from the lowest CATE scores. Contrary to `prop.cutoff`, `prop.multi` must include both 0 and 1 to remind us of the mutual exclusiveness. **precmed** will automatically attach 0 and/or 1 if they are not supplied and a warning message will be given. 

Now we show more examples of different cutoffs and how they are reflected in the results.

```{r subgroup_proportion, eval = T, echo = T, cache = doc.cache}
# An example of 9 nested binary subgroups and a 4-category mutually exclusive subgroup
tau0 <- with(survivalExample,
             min(quantile(y[trt == "drug1"], 0.95), quantile(y[trt == "drug0"], 0.95)))

output_catecv2 <- catecv(response = "survival",
                         data = survivalExample,
                         score.method = c("randomForest", "contrastReg"),
                         
                cate.model = survival::Surv(y, d) ~ age + female
                                                        + previous_cost + previous_number_relapses,
                ps.model = trt ~ age + previous_treatment,
                initial.predictor.method = "logistic",
                ipcw.model = NULL,
                followup.time = NULL,
                tau0 = tau0,
                surv.min = 0.025,
                higher.y = TRUE,
                prop.cutoff = seq(0.6, 1, length = 9), # NEW
                prop.multi = c(0, 0.4, 0.5, 0.6, 1),   # NEW
                cv.n = 5,
                seed = 999,
                plot.gbmperf = FALSE,
                verbose = 0)
```

```{r subgroup_proportion2, eval = T, echo = T}
print(output_catecv2$ate.contrastReg$ate.est.train.high.cv) 
# Dimension is 9 (nested binary subgroups) by 5 (CV iterations)
```

Because we have new `prop.cutoff` with 9 values as opposed to only 5 in the original example, the number of rows in the ATE results will be changed accordingly while the number of columns remains to be 5.

```{r subgroup_proportion3, eval = T, echo = T}
plot(output_catecv2) 
```

The validation curves depend on `prop.cutoff` and have more line segments than the one presented in the main example because they contain 9 sets of subgroups instead of 5 sets (although it is not so obvious here in this example). The ABC statistic is not affected by this because the calculation of ABC statistics depends on the minimum and maximum of
the `prop.cutoff` vector and 100 evenly-spaced subgroup cutoffs are created from this range. The highest proportion cutoff that is not 1 in the main example was 0.9 and it is now 0.95. See more theoretical details on ABC calculation in the [Theoretical details](Theoretical-details.html) section.

```{r subgroup_proportion4, eval = T, echo = T}
print(output_catecv2$ate.contrastReg$ate.est.train.group.cv) 
# Dimension is now 4 (multi-category mutually exclusive subgroups) by 3 (CV iterations)

boxplot(output_catecv2)
```

For results on the mutually exclusive subgroups, the number of rows in the ATE results increases from 3 to 4 in comparison with the original example because `prop.multi` now defines 4 mutually exclusive subgroups. For each methods, the box plot now has 4 boxes for 4 mutually exclusive subgroups.

**Examples that we do not recommend**

In this example, `prop.cutoff` includes 0.01, which is very close to 0. Consequently, the scoring methods may run into some convergence issues because the first subgroup includes a small number of patients. This can lead to highly unstable ATE estimates, as shown in the first CV iteration of contrast regression in `output_catecv3$ate.contrastReg$ate.est.valid.high.cv`. Moreover, the validation curves become useless due to the extreme ATE estimates.

```{r subgroup_proportion_bad, eval = T, echo = T, error = T, cache = doc.cache}
# An example of very few nested binary subgroups
output_catecv3 <- catecv(response = "survival",
                         data = survivalExample,
                         score.method = c("contrastReg"),
                         cate.model = survival::Surv(y, d) ~ age + female + previous_cost + previous_number_relapses,
                         ps.model = trt ~ age + previous_treatment,
                         initial.predictor.method = "logistic",
                         ipcw.model = NULL,
                         followup.time = NULL,
                         tau0 = tau0,
                         surv.min = 0.025,
                         higher.y = TRUE,
                         prop.cutoff = c(0, 0.1, 0.75, 1), # NEW
                         prop.multi = c(0, 0.5, 0.6, 1),
                         cv.n = 5,
                         seed = 999,
                         plot.gbmperf = FALSE,
                         verbose = 1)
output_catecv3$ate.contrastReg$ate.est.valid.high.cv

```

```{r subgroup_proportion_bad2, eval = T, echo = T, warning=FALSE, message=FALSE}
plot(output_catecv3) 
```

Below are some general recommendations on the choice of `prop.cutoff` and `prop.multi`:

1.  We do not recommend including 0 in `prop.cutoff` as it is an invalid value. **precmed** automatically removes 0 in `prop.cutoff` and outputs a warning message.
2.  We do not recommend choosing values too close to 0 either because this will generate very small subgroups which may lead to numerical instability, e.g., 10/90 split specified by 0.1 in this example. In this situation, either the validation curves will be plotted by the range of the unstable ATE estimates will make them useless or the validation curves will not be plotted if the numerical instability due to an extreme split leads to missing or infinite values, in which case a warning or error is printed as in the example above.
3.  In general, we do not recommend specifying too few proportions for `prop.cutoff` (e.g., 2 or 3) because the validation curves will look very jagged. The validation curves will be smoother with more proportions and more spread out over the range of proportions. If computation is an issue, however, choosing too many values for `prop.cutoff` or `prop.multi` can be time consuming. It is recommended to start from a smaller length and you can increase the length as needed.

## IPCW model and method

The default IPCW model uses the same covariates as the outcome model `cate.model` plus the treatment. The default IPCW method is a Cox regression with Breslow estimator of the baseline survivor function from `coxph()` function in the `survival` package. Different variables or different method can be specified in the IPCW model. In the example below, we changed the covariates to the number of symptoms and number relapses in the pre-index period and specified the Weibull accelerated failure time (AFT) regression model using the location-scale parameterization. The `survreg()` function in the `survival` package was used to fit the AFT model.

```{r ipcw, eval = F, echo = T, cache = doc.cache}
# An example of a different IPCW model with different covariates from default
output_catecv4 <- catecv(response = "survival",
                         data = survivalExample,
                         score.method = c("poisson", "contrastReg"),
                         cate.model = survival::Surv(y, d) ~ age + female + previous_cost + previous_number_relapses,
                         ps.model = trt ~ age + previous_treatment,
                         initial.predictor.method = "logistic",
                         ipcw.model = ~ previous_number_symptoms + previous_number_relapses, # NEW
                         ipcw.method = "aft (weibull)", # NEW
                         followup.time = NULL,
                         tau0 = tau0,
                         surv.min = 0.025,
                         higher.y = TRUE,
                         prop.cutoff = seq(0.6, 1, length = 5), 
                         prop.multi = c(0, 0.5, 0.6, 1),
                         cv.n = 5,
                         seed = 999,
                         plot.gbmperf = FALSE,
                         verbose = 0)
```

## Truncation and follow-up time

The argument `tau0` specifies the truncation time for RMTL. The default is the maximum survival time in the data but it can be changed if you have a specific truncation time in mind. Similarly, the maximum follow-up time argument `followup.time` can be changed from the default, which is unknown potential censoring time. The study setup usually can shed some light on what values to use for these two time parameters.

In the main example, we used the minimum of the 95th percentile of survival time in either treatment group as `tau0`. Below we show how setting different values of the truncation time.

```{r time, eval = T, echo = T, cache = doc.cache}
# An example of a different IPCW model with different covariates from default
output_catecv5 <- catecv(response = "survival",
                         data = survivalExample,
                         score.method = c("poisson", "contrastReg"),
                         cate.model = survival::Surv(y, d) ~ age + female + previous_cost + previous_number_relapses,
                         ps.model = trt ~ age + previous_treatment,
                         initial.predictor.method = "logistic",
                         ipcw.model = ~ previous_number_symptoms + previous_number_relapses,
                         ipcw.method = "aft (weibull)",
                         followup.time = NULL,
                         tau0 = NULL, # NEW
                         higher.y = TRUE,
                         cv.n = 5,
                         surv.min = 0.025,
                         prop.cutoff = seq(0.6, 1, length = 5),
                         prop.multi = c(0, 0.5, 0.6, 1),
                         seed = 999,
                         plot.gbmperf = FALSE,
                         verbose = 0)
plot(output_catecv5)
```


# Other `precmed` vignettes in this serie  
  
[1. Examples for count outcome](Count-examples.html)  
[2. Examples for survival outcome](Survival-examples.html)  
[3. Additional examples](Additional-examples.html)  
[4. Theoretical details](Theoretical-details.html)    
    
