---
title: "precmed Overview"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Software

1. [`precmed` package at CRAN](https://cran.r-project.org/package=precmed)
2. [precmed GitHUB repository](https://github.com/smartdata-analysis-and-statistics/precmed)

## Installation

The `precmed` package can be installed from CRAN as follows:

```{r eval = FALSE}
install.packages("precmed")
```

The latest version can be installed from GitHub as follows: 

```{r eval = FALSE}
install.packages("devtools")
devtools::install_github("smartdata-analysis-and-statistics/precmed")
```

## Capabilities of `precmed` package

The `precmed` package contains functions to

- Estimate average treatment effects for count, survival and contiuous data
- Estimate conditional average treatment effect (CATE) score for count, survival and continuous data
- Cross-validation of CATE score
- Compute area between curves
- Plot validation curves


## Main functions

The main functions in the `precmed` package are:

Function name      | Description
-------------------|---------------------------------
atefit()           | Doubly robust estimator of and inference for the average treatment effect for count, survival and continuous data
catefit()          | Estimation of the conditional average treatment effect (CATE) score for count, survival and continuous data
catecv()           | Cross-validation of the conditional average treatment effect (CATE) score for count, survival or continuous outcomes
abc()              | Compute the area between curves from the `precmed` object
plot()             | Two side-by-side line plots of validation curves from the `precmed` object
boxplot()          | A set of box plots of estimated ATEs from the `precmed` object

## Background materials

1. [Zhao et al. (2013), Effectively selecting a target population for a future comparative study](https://doi.org/10.1080/01621459.2013.770705)

2. [Yadlowsky et al. (2020), Estimation and validation of ratio-based conditional average treatment effects using observational data](https://doi.org/10.1080/01621459.2020.1772080)


## Vignettes

Under construction

1. [General introduction to precmed]
2. [Example for count outcome]
3. [Example for survival outcome]
4. [Detailed examples]
5. [Theoretical details]


## References
